---
title: "Lab 11 - Stats 503"
output:
  html_document: default
---

PITCHf/x data from MLB. Contains variables:

* `start_speed`: Initial speed in mph
* `pfx_x`, `pfx_z`: Horizontal and vertical movement relative to a ball thrown
with zero spin.
* `x0`, `z0`: Horizontal and vertical location of the release point of the pitch.

```{r, message=F, warning=F}
require(ggplot2)
require(reshape2)
require(gridExtra)

# require(pitchRx)
# require(plyr)
# rawdat = scrape(start = "2016-10-01", end = "2016-10-01")
# vars_pitch = c('start_speed', 'pfx_x', 'pfx_z', 'x0', 'z0', 'num', 'url')
# vars_atbat = c('num', 'url', 'p_throws')

# dat = join(rawdat$atbat[,vars_atbat], rawdat$pitch[,vars_pitch],
#             by=c('num', 'url'), type='inner')
# dat = dat[complete.cases(dat),]
# dat = dat[1:1500,]

# keepvars = c('start_speed', 'pfx_x', 'pfx_z', 'x0', 'z0')
# write.csv(dat[,keepvars], file='./data/pitches.txt', row.names=FALSE)

dat = read.csv('./data/pitches.txt')

```


## Visualization

```{r, message=F, warning=F}
dat_norm = scale(dat)
dis = dist(dat_norm)
mds = as.data.frame(cmdscale(dis, k=2))

q1 = ggplot(mds, aes(x=V2, y=V1)) + geom_point(alpha=0.6)
q2 = ggplot(dat, aes(x=pfx_x, y=pfx_z)) + geom_point(alpha=0.6)
q3 = ggplot(dat, aes(x=x0, y=z0)) + geom_point(alpha=0.6)
q4 = ggplot(dat, aes(x=pfx_x, y=start_speed)) + geom_point(alpha=0.6)
grid.arrange(q1, q2, q3, q4, ncol=2)
```


## Hierarchical Clustering

Packages:

* `hclust` in base R.
* `agnes` in the `cluster` package.

```{r, message=F, warning=F}
require(cluster)

pitch_sing = agnes(dat_norm, diss=FALSE, method='single')
pitch_comp = agnes(dat_norm, diss=FALSE, method='complete')
pitch_ward = agnes(dat_norm, diss=FALSE, method='ward')

sil_sing = silhouette(cutree(pitch_sing, k=5), dis)
sil_comp = silhouette(cutree(pitch_comp, k=5), dis)
sil_ward = silhouette(cutree(pitch_ward, k=5), dis)

```

```{r, message=F, warning=F}
require(ggdendro)

# plot(pitch_sing)

# ggdendrogram(as.dendrogram(pitch_sing), leaf_labels=FALSE, labels=FALSE)
# summary(sil_sing)
# plot(sil_sing)

# ggdendrogram(as.dendrogram(pitch_comp), leaf_labels=FALSE, labels=FALSE)
# summary(sil_comp)
# plot(sil_comp)

ggdendrogram(as.dendrogram(pitch_ward), leaf_labels=FALSE, labels=FALSE)
summary(sil_ward)
plot(sil_ward)

```


## Exploring numbers of clusters

https://rogerfan.shinyapps.io/lab11-shiny/


## K-Means

We can use the built-in function `kmeans` to do k-means clustering. The basic
algorithm can be selected with the option `algorithm='Lloyd'`, but the default
algorithm has several optimizations that result in better performance.

K-means is very sensitive to initial values, so as a rule always use multiple
starts and choose the best one.

```{r, message=F, warning=F}
# ?kmeans
set.seed(23462)

clust_kmeans = kmeans(dat_norm, 5, iter.max=100, algorithm='Lloyd')
clust_kmeans = kmeans(dat_norm, 5)
clust_kmeans

```

```{r, message=F, warning=F}
set.seed(625723)  # For a bad example
clust_kmeans2 = kmeans(dat_norm, 5)

mds_temp = cbind(
    mds, as.factor(clust_kmeans$cluster), as.factor(clust_kmeans2$cluster))
names(mds_temp) = c('V1', 'V2', 'clust1', 'clust2')

gp1 = ggplot(mds_temp, aes(x=V2, y=V1, color=clust1)) +
    geom_point() + theme(legend.position="none")
gp2 = ggplot(mds_temp, aes(x=V2, y=V1, color=clust2)) +
    geom_point() + theme(legend.position="none")
grid.arrange(gp1, gp2, ncol=2)

clust_kmeans$tot.withinss
clust_kmeans2$tot.withinss
clust_kmeans$betweenss
clust_kmeans2$betweenss

```


## K-Means++

A smarter way of initializing K-Means that usually results in better
performance. An implementation is available in the `flexclust` package, but the
basic algorithm is easy to implement.

The algorithm is:

* For each data point, compute the distance to the nearest existing center.
* Choose a new data point at random to be a center, weighting the draw
so the probability that a point is chose is proportional to its squared distance.
* Repeat until K centers are chosen.
* Apply K-means with those centers as the initial ones.

This usually performs better and more consistently than K-means, but you should
still use multiple initializations.

```{r, message=F, warning=F}
require(matrixStats)

kpp_init = function(dat, K) {
    x = as.matrix(dat)
    n = nrow(x)

    # Randomly choose a first center
    centers = matrix(NA, nrow=K, ncol=ncol(x))
    centers[1,] = as.matrix(x[sample(1:n, 1),])

    for (k in 2:K) {
        # Calculate dist^2 to closest center for each point
        dists = matrix(NA, nrow=n, ncol=k-1)
        for (j in 1:(k-1)) {
            temp = sweep(x, 2, centers[j,], '-')
            dists[,j] = rowSums(temp^2)
        }
        dists = rowMins(dists)

        # Draw next center with probability propor to dist^2
        cumdists = cumsum(dists)
        prop = runif(1, min=0, max=cumdists[n])
        centers[k,] = as.matrix(x[min(which(cumdists > prop)),])
    }
    return(centers)
}

clust_kpp = kmeans(dat_norm, kpp_init(dat_norm, 5), iter.max=100, algorithm='Lloyd')

```


## Exercise: Gaussian Mixture Models

* Generate a 2-d dataset with 50 points from a $N((0, 0), 6 \times I)$ distribution,
50 from a $N((4, 4), 0.1 \times I)$, and 50 from a $N((-4, 0), 0.1 \times I)$.
* Fit a Gaussian mixture model on the data and generate cluster labels.
To do this, investigate the `mvnormalmixEM` function in the package `mixtools`.
Keep in mind that GMM can be sensitive to initial values.
To generate the labels, use `$posterior` from the result.
* Fit a K-means model on the data and generate cluster labels.
* Plot the results from both models side-by-side. Comment on the results.

```{r, message=F, warning=F, eval=F, echo=F}
require("mixtools")
set.seed(28748)

n_each = 50
x1 = matrix(rnorm(n_each*2, sd=sqrt(6)), ncol=2)
x2 = sweep(matrix(rnorm(n_each*2, sd=sqrt(0.1)), ncol=2), 1, c(4, 4), '+')
x3 = sweep(matrix(rnorm(n_each*2, sd=sqrt(0.1)), ncol=2), 2, c(-4, 4), '+')

x = as.data.frame(rbind(x1, x2, x3))
ggplot(x, aes(x=V1, y=V2)) + geom_point()
```

```{r, message=F, warning=F, eval=F, echo=F}
gmm = mvnormalmixEM(as.matrix(x), k=3)
kmeans = kmeans(x, 3)

x$gmm = as.factor(apply(gmm$posterior, 1, which.max))
x$kmeans = as.factor(kmeans$cluster)
x_melt = melt(x, id.vars=c('V1', 'V2'))

ggplot(x_melt, aes(x=V1, y=V2, color=value)) +
    geom_point() + facet_grid(~variable) + theme(legend.position='none')

```



