---
title: "Lab 2 - Stats 503"
output:
  html_document: default
  html_notebook: default
---

## Functional programming

Download from Canvas the data file "citycrime.txt", which shows crime rates per 100,000 people on some US cities. 
```{r}
data_crime = read.table("citycrime.txt", header = TRUE)
```

The functions <tt>sapply</tt>,  <tt>lapply</tt>, <tt>apply</tt> and <tt>vapply</tt> can apply a function to each element of an object, but the output and syntax is different. These functions are called *functionals*.

```{r, eval=FALSE}
colMeans(data_crime)
sapply(data_crime, mean)
lapply(data_crime, mean)
apply(data_crime, 2, mean)
vapply(data_crime, mean, double(1))

```

Examples with user-defined functions to calculate descriptive statistics.
```{r}
summary_stats = function(x) {
  return(c(mean = mean(x),
              sd = sd(x),
              max = max(x),
              min = min(x),
              med = median(x)))
}

crime_stats = lapply(data_crime, summary_stats)
```
Calculate the cities with higher crime rates.
```{r}
cities = rownames(data_crime)

max_city = function(x) {
  cities[which.max(x)]
}
sapply(data_crime, max_city)
```

It is very easy to implement parallel programming using functionals.
```{r, eval=FALSE}
library(parallel)
wait = function(x) {
   mean(x)
   Sys.sleep(20)
}
num_cores = detectCores() - 1
cl = makeCluster(num_cores)
parLapply(cl, data_crime, wait)
stopCluster(cl)
```

The <tt>Map</tt> function is similar to <tt>lapply</tt> but it apply the function to the elements of multiple arguments. An example to calculate a weighted mean.
```{r, eval=FALSE}
cost_per_city = replicate(7, runif(length(cities)), simplify = FALSE) 
Map(weighted.mean, data_crime, cost_per_city)
```

<tt>Reduce</tt> succesively combines the elements of a a vector. <tt>Filter</tt> applies a logical function to the elements of a list and returns those that satisfy the condition. There are more functionals for different purposes (<tt>Position</tt>, <tt>Find</tt>, etc.).

```{r}
top10_cities = lapply(data_crime, function(x) cities[order(x, decreasing = T)[1:10]])

intersect(intersect(intersect(top10_cities$Murder, top10_cities$Rape), top10_cities$Robbery), top10_cities$Assault) #... and so on

Reduce(intersect, top10_cities)

Filter(function(x) "Detroit" %in% x, top10_cities)
```


### Reshaping a dataset

There are two formats that a dataset can be presented: long and wide. The functions <tt>melt</tt> and <tt>dcast</tt> can modify from one format to another.

```{r}
library(reshape2)
data_crime_city = data.frame(data_crime, City = rownames(data_crime))
head(data_crime_city)
long_data_crime = melt(data_crime_city, id.vars = "City")
head(long_data_crime)
wide_data_crime = dcast(long_data_crime, City ~ variable)
head(wide_data_crime)
```

### Functionals with data frames
The library <tt>plyr</tt> contains functions to perform functional programming. In particular, it is useful to manipulate data frames.
```{r, eval=FALSE}
library(plyr)
summarize(data_crime, mean_murder = mean(Murder), mean_robbery = mean(Robbery))
transform(data_crime, murder_square = Murder^2, mean_robbery = mean(Robbery))
```
The function <tt>ddply</tt> is similar to <tt>apply</tt>, but it takes as input and output a data.frame.
```{r, eval=FALSE}
ddply(long_data_crime, c("City"), function(df) mean(df$value))
ddply(long_data_crime, c("City"), summarize, mean_city =mean(value), sd_city = sd(value))
ddply(long_data_crime, c("City"), summarize, mean_city =mean(value), sd_city = sd(value))
```

## Visualization using ggplot
The base plotting functions in R are useful to produce simple outputs, but as the level of difficulty in the plot progresses, they can become difficult to modify. The library <tt>ggplot2</tt> offers a plotting system with substantial capabilities and nicer graphics than the base plotting functions.
```{r}
plot(x = data_crime$Robbery, y = data_crime$Assault)
```

The function <tt>qplot</tt> (quick plot) is an equivalent to <tt>plot</tt>. Useful to create simple plots, but it has a limited functionality.
```{r}
library(ggplot2)
qplot(x=Robbery, data = data_crime, geom="histogram")
qplot(x = Robbery, y = Assault, data = data_crime)
```

The <tt>ggplot()</tt> is the standard way to create plots. 
```{r}
crime_plot = ggplot(data_crime, aes(x = Robbery, y = Assault))
crime_plot + geom_point()
```

```{r}
crime_plot + geom_point(aes(color = Murder, size = Larceny))
```

```{r}
crime_plot + geom_point(aes(color = Murder)) + 
  scale_colour_gradient(high = "red", low = "blue") + theme_bw()

```


```{r}
crime_plot + geom_text(aes(label=rownames(data_crime)), check_overlap = TRUE)
```

Sometimes it is useful to create 3D plots to visualize multivariate data. The library <tt>plotly</tt>  can do so.
```{r,warning=FALSE}
#install.packages("plotly")
library(plotly)
p = plot_ly(data_crime, x = ~Robbery, y = ~Assault, z = ~Murder)
p
```

Pairs plots are useful to visualize multiple variables at the same time. The base function is <tt>pairs</tt>, but here is an example with ggplot.
```{r}
library(GGally)
crime_pairs = ggpairs(data_crime,axisLabels = "none",
        upper = list(continuous = "points", combo = "dot"),
        lower = list(continuous = "cor", combo = "dot"),
        diag = list(continuous = "densityDiag")) + 
  theme_bw()
crime_pairs
```



Ggplot easily allows to visualize multiple time series in the same plot. For this, it is necessary to convert the data first to a long format using <tt>melt</tt>.
```{r}
trigon_data = data.frame(lapply(c(sin, cos), function(f) f(seq(0, 2*pi,0.05))), x= seq(0, 2*pi,0.05))
colnames(trigon_data) = c("sin", "cos", "x")
trigon_data_long = melt(trigon_data, id.vars = "x")
ggplot(trigon_data_long, aes(x=x, y=value,colour=variable, group = variable)) + geom_line()
```


Finally, R has functions that allow to save a plot to a file. Different extensions are allowed, here is an example with <tt>pdf</tt>.
```{r}
pdf("crime_ggpairs.pdf")
crime_pairs
dev.off()
```

## Exercises

Work in pairs to solve the next questions. In all problems, try to use functional programming when possible, and ggplot.

1. Simulate data from a multivariate normal distribution wih mean 0 and covariance diag($3^2, 2^2,1,1,\ldots$) using p=20 variables and n=100 observations. Compute the eigenvalues of the sample covariance and plot them.

2. Create a list containing different versions of the simulated data varying the number of observations on each one, from 100 to 2000. For each element in the list, compute the SVD and calculate the running time. Store the results on an array. Plot the running times as a function of the number of observations. (You can use proc.time()).

3. Repeat the previous problem, but using the eigendecomposition of the covariance matrix.

4. Repeat 2 and 3, but instead of varying the number of observations, change the number of variables from 100 to 2000.

```{r, echo=FALSE}
sample = matrix(rnorm(1000*10), ncol = 10)
p = 20
n = 100
X = sweep(matrix(rnorm(n*p),ncol = p), 2, c(3,2,rep(1,p-2)), "*")
eigenvalues = eigen(cov(X))$values
qplot(y=eigenvalues, xlab="Index", ylab="Eigenvalues")
```





```{r, echo=FALSE}
generate_data = function(n, p) {
  X = sweep(matrix(rnorm(n*p),ncol = p), 2, c(3,2,rep(1,p-2)), "*")
}

compute_svd = function(X) {
  start = proc.time()
  X = scale(X, center = TRUE, scale = FALSE)
  eigenvalues = svd(X)$d^2/(nrow(X)-1)
  end = proc.time()
  return(end[3] - start[3])
}

data_obs = lapply(seq(100, 5000, 100), generate_data, p = 100)
svd_times = sapply(data_obs, compute_svd) 
qplot(y = svd_times, main = "SVD", xlab = "Observations", ylab = "Running time")

```


```{r,cache=TRUE, echo=FALSE}
compute_eig = function(X) {
  start = proc.time()
  eigenvalues = eigen(cov(X))$values
  end = proc.time()
  return(end[3] - start[3])
}
eig_times = sapply(data_obs, compute_eig) 
qplot(y = eig_times, main = "Eig", xlab = "Observations", ylab = "Running time")
```

```{r,cache=TRUE,echo=FALSE}
data_var = lapply(seq(100, 2000, 100), generate_data, n = 100)
svd_times = sapply(data_var, compute_svd) 
qplot(y = svd_times, main = "SVD", xlab = "Variables", ylab = "Running time")
eig_times = sapply(data_var, compute_eig) 
qplot(y = eig_times, main = "Eig", xlab = "Variables", ylab = "Running time")
```

