---
title: "Lab 3 - Stats 503"
output:
  html_document: default
  html_notebook: default
---

# Read in data

```{r, message=F, warning=F}
data = read.csv('./data/cars.csv')
head(data)
```


## Calculate PCA

```{r, message=F, warning=F}
X = data[,-1]
pca = princomp(X, cor=FALSE)
summary(pca)
loadings(pca)


pca$sdev^2

covX = cov(X)
eig = eigen(covX)
eigen(covX)$values

X_demean = scale(X, center=TRUE, scale=FALSE)
svd(X_demean)$d^2/(nrow(X)-1)
```


## Scree Plot

```{r, message=F, warning=F}
require(ggplot2)

plot(pca)

scree = qplot(1:length(pca$sdev), pca$sdev, geom='line',
              ylab='Component Standard Deviation', xlab='Component')
scree
```


## Biplot

```{r, message=F, warning=F}
# library(devtools)
# install_github("vqv/ggbiplot")

require(ggbiplot)

ggbiplot(pca)
```


## PCA as Projection

```{r, message=F, warning=F}
require(plotly)

data_res = data[,c('mpg', 'hp', 'wt')]

plot_data = plot_ly(data_res, x = ~mpg, y = ~hp, z = ~wt)
plot_data

X_res = data[,c('mpg', 'hp', 'wt')]
pca_res = princomp(X_res, cor=FALSE)

W = loadings(pca_res)[,1:2]

proj = pca_res$scores[,1:2] %*% t(W)
proj = sweep(proj, 2, pca_res$center, "+")

proj = as.data.frame(proj)
proj$lab = 'Projected'
X_res$lab = 'Original'
data_for_plotting = rbind(X_res, proj)

plot_pca1 = plot_ly(data_for_plotting, x=~mpg, y=~hp, z=~wt, color=~lab)
plot_pca1
```


```{r, message=F, warning=F}
require(scatterplot3d)

data_for_plotting$col = ifelse(data_for_plotting$lab == 'Original', 'blue', 'red')

scatterplot3d(
    data_for_plotting[,1:3], color=data_for_plotting$col,
    angle=160, scale.y=0.6,
    pch=19,type='h', lty.hplot=2)
```


## Bootstrap

```{r, message=F, warning=F}
set.seed(2346234)
num_boot = 1000

ols = lm(mpg ~ disp + hp + drat + wt + qsec, data=data)
coef(ols)['wt']

bootstrap_iter = function(i) {
    data_resample = data[sample(1:nrow(data), nrow(data), replace=TRUE),]
    ols_resample = lm(mpg ~ disp + hp + drat + wt + qsec, data=data_resample)
    return(coef(ols_resample)['wt'])
}

bootstrap_res = data.frame(coef=sapply(1:num_boot, bootstrap_iter))

plot_bootstrap = ggplot(bootstrap_res, aes(x=coef)) +
    geom_histogram() + geom_vline(xintercept=coef(ols)['wt'], color='magenta')
plot_bootstrap
```


## Exercises

1. Generate $X$ as 200 observations from a $10$-dimensional multivariate
normal distribution with a mean $\mu = (1, 2, \dots, p)$ and an AR1(0.6) covariance
matrix, i.e. $\Sigma_{ij} = 0.6^{\lvert i - j \rvert}$.
2. Calculate the largest singular value of $X$.
3. Use the bootstrap to estimate the sampling distribution of the largest
singular value of $X$.
4. Repeat steps 1-2 many times to directly simulate the sampling distribution.
5. Plot the simulated and bootstrapped distributions on the same plot and
compare.

```{r, message=F, warning=F}
# Part 1

set.seed(56256)

rho = 0.6
n = 100
p = 10

Sigma = diag(p)
Sigma = rho^abs(row(Sigma)-col(Sigma))

X = matrix(rnorm(n*p), ncol=p) %*% chol(Sigma)
X = sweep(X, 2, 1:p, "+")
```

```{r, message=F, warning=F}
# Part 2

svd(X)$d[1]
```

```{r, message=F, warning=F}
# Part 3

num_boot = 500

bootstrap_iter = function(i) {
    X_resample = X[sample(1:n, n, replace=TRUE),]
    return(svd(X_resample)$d[1])
}
res_boot = data.frame(d=sapply(1:num_boot, bootstrap_iter))
```

```{r, message=F, warning=F}
# Part 4

res_sim = data.frame(d=rep(NA, num_boot))
for (i in 1:num_boot) {
    X_sim = matrix(rnorm(n*p), ncol=p) %*% chol(Sigma)
    X_sim = sweep(X_sim, 2, 1:p, "+")
    res_sim$d[i] = svd(X_sim)$d[1]
}
```

```{r, message=F, warning=F}
# Part 5

res_boot$labels = 'Bootstrap'
res_sim$labels  = 'Simulated'
res_all = rbind(res_boot, res_sim)

values = data.frame(x=c(eigen(Sigma)$values[1], svd(X)))

plot_dists = ggplot(res_all, aes(x=d, fill=labels)) +
    geom_histogram(alpha=0.5, aes(y=..density..), position='identity') +
    ylab('Density') + xlab('Largest Singular Value') +
    theme_bw() +
    theme(legend.position=c(0.99,0.99), legend.justification=c(1,1),
          legend.title=element_blank(),
          legend.background=element_rect(color="lightgrey"))

# pdf('./sampling_dists.pdf', height=6, width=7.5)
# print(plot_dists)
# dev.off()
plot_dists
```


