---
title: "Lab 5 - Stats 503"
output:
  html_document: default
  html_notebook: default
---


Any time you have issues, your first reference should be the ggplot2
help page at [http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/).


Building blocks of a ggplot graph:

* Data
* Aesthetic mapping
* Geometric objects
* Scales and legends
* Faceting and visuals

```{r, message=F, warning=F}
#install.packages('ISLR')
require(ISLR)
require(ggplot2)
require(reshape2)

data(Carseats)
head(Carseats)

data(EuStockMarkets)
stock_data = as.data.frame(EuStockMarkets)
stock_data$time = as.numeric(time(EuStockMarkets))
head(stock_data)
```


## Aesthetic mapping

How do we map variables to features of the plot. Common examples:

* Position (x, y coordinates)
* Color
* Fill ('inside' coloring)
* Shape
* Linetype
* Size

```{r, message=F, warning=F}
plot1 = ggplot(data=Carseats, aes(x=Price, y=Sales, color=ShelveLoc, shape=Urban)) +
    geom_point()
plot1

ggplot(data=Carseats, aes(x=Price, y=Sales, color=Advertising)) +
    geom_point() +
    xlab('Prices ($)') + ylab('Sales (thousands)') + ggtitle('Car Seat Sales')
```


Note that aesthetic mappings are inherited from the default aesthetic but can
also be applied for specific pieces.

```{r, message=F, warning=F}
ggplot(data=Carseats, aes(x=Price, y=Sales)) +
    geom_point(aes(color=US))
```


Similarly, data can also be applied
for a specific object, but otherwise is inherited from the top-level.

```{r, message=F, warning=F}
ggplot(data=Carseats, aes(x=Price, y=Sales)) +
    geom_point(data=subset(Carseats, US == 'Yes'), aes(color=Urban)) +
    geom_point(data=subset(Carseats, US == 'No'))
```


## Geometric objects

What marks do we want on the plot. Each `geom_` object tends to have
certain `aes` arguments that it requires or can use. Examples:

* `geom_point` for scatter plots
* `geom_line` for line plots
* `geom_histogram`
* `geom_text`

When using multiple geometries, you can direcly modify their attributes
or add them to the aesthetic (so they get included in legends).

```{r, message=F, warning=F}
Carseats_addpred = Carseats
Carseats_addpred$Sales_pred = predict(lm(Sales ~ Price, data=Carseats))

ggplot(data=Carseats_addpred, aes(x=Price)) +
    geom_point(aes(y=Sales)) +
    geom_line(aes(y=Sales_pred), color='Magenta')

ggplot(data=Carseats_addpred, aes(x=Price)) +
    geom_point(aes(y=Sales, color='Data')) +
    geom_line(aes(y=Sales_pred, color='Predicted'))
```


Some miscellaneous other `geom_` examples.

```{r, message=F, warning=F}
ggplot(data=stock_data[1:100,], aes(x=time, y=DAX)) +
    geom_point() +
    geom_smooth(span=0.3)

ggplot(data=Carseats, aes(x=Price, y=Sales)) +
    geom_point() +
    geom_smooth(method='lm', se=FALSE)

ggplot(data=Carseats, aes(x=Sales, color=ShelveLoc, fill=ShelveLoc)) +
    geom_density(alpha=0.3)

ggplot(data=Carseats, aes(x=Price, y=Sales)) +
    geom_point() +
    geom_abline(intercept=10, slope=-0.04) +
    geom_vline(xintercept=115, linetype='dashed')

require(mvtnorm)
dat = Carseats[,c('Price', 'Sales')]
xgrid = expand.grid(
    Price=seq(24, 191, length.out=50),
    Sales=seq(0, 16.3, length.out=50))
xgrid$p = dmvnorm(xgrid, mean=colMeans(dat), sigma=cov(dat))

ggplot(Carseats, aes(x=Price, y=Sales)) +
    geom_point() +
    geom_contour(aes(z=p), data=xgrid, breaks=3e-4)

ggplot(Carseats, aes(x=Price, y=Sales)) +
    geom_point() +
    stat_ellipse()
```


## Aside on melt

You'll very often find that you need data in long format to plot it using
ggplot, which means you will want to use `melt` from the reshape2 package.
To change data in the other direction, from long to wide, you can use `dcast`.

```{r, message=F, warning=F}
head(stock_data)

stock_data_melted = melt(stock_data, id.vars='time')
head(stock_data_melted)

ggplot(data=stock_data_melted, aes(x=time, y=value, color=variable)) +
    geom_line()
```


## Scales and Legends

```{r, message=F, warning=F}
plot2 = ggplot(data=Carseats, aes(x=Price, y=Sales, color=Advertising, shape=ShelveLoc)) +
    geom_point()

plot2 + scale_shape_discrete(name='Shelve Location',
                             breaks=c('Good', 'Medium', 'Bad'),
                             labels=c('G', 'M', 'B'))

plot2 + scale_color_continuous(name='Advertising Level',
                               breaks=seq(0, 30, 5),
                               low='grey', high='red')

require(RColorBrewer)
plot2 + scale_color_distiller(palette='YlOrRd', direction=1)

plot2 + scale_x_continuous(limits=c(0, 300))

plot2 + scale_y_log10()

plot2 + scale_y_reverse()

plot2 + coord_fixed(ratio=5)
```


## Faceting and visuals

```{r, message=F, warning=F}
plot2 + facet_grid(. ~ Urban)

plot2 + facet_grid(US ~ Urban, labeller=label_both)

ggplot(data=stock_data_melted, aes(x=time, y=value)) +
    geom_line() +
    facet_wrap(~variable)

```

## Themes

```{r, message=F, warning=F}
plot2 + theme_bw()

plot2 + theme_minimal()

plot2 + theme(legend.position=c(0.98,0.98), legend.justification=c(1,1),
              legend.box = "horizontal",
              legend.background=element_rect(color="lightgrey"))
```


## Exercises

1. Plot the centers of each US state by their latitude and longitude. Label the
states with their two-letter abbreviation. Size the dots according to each state's
population. Color the states according to their regional division.
Note that you can load this data using `data(state)`, after which state data will be
located in:

    * `state.center`
    * `state.abb`
    * `as.data.frame(state.x77)$Population`
    * `state.division`

```{r, message=F, warning=F}
data(state)
data_state = data.frame(
    x=state.center$x, y=state.center$y, abb=state.abb,
    Population=as.data.frame(state.x77)$Population, Division=state.division)

ggplot(data_state, aes(x=x, y=y)) +
    geom_point(aes(size=Population, fill=Division),
               pch=21, color='black', stroke=0.2, alpha=0.7) +
    geom_text(aes(label=abb), size=2, hjust=0.5, vjust=-0.5) +
    scale_size_area() +
    scale_y_continuous(limits=c(15, 60)) +
    coord_fixed(ratio=1)
```


2. `dowjones.csv` contains weekly percentage changes for a set of 30 stocks.
Attempt to replicate the following figure, which visualizes the correlation
matrix of the stocks.

    Hints:

    * You might want to look up `geom_tile`.
    * Try using `melt` on the correlation matrix.
    * If you want a "prettier" ordering of variables, you can use
    `hclust(distance_matrix)$order` to extract one based on hierarchical clustering,
    as long as you can get an appropriate distance matrix. You might need to use `as.dist`
    to convert a matrix to a "distance matrix" form.

```{r, message=F, warning=F}
# Read in data
data = read.csv('./data/dowjones.csv')
data_wide = dcast(data, date ~ symb)
corr_mat = cor(data_wide[,-1])

# Use hierarchical clustering to order
dist_temp = as.dist(1-corr_mat)
hc = hclust(dist_temp)
corr_mat = corr_mat[hc$order, hc$order]

# Remove upper triangle
corr_mat_lower = corr_mat
corr_mat_lower[upper.tri(corr_mat_lower)] = NA

# Melt correlation matrix and make sure order of factor variables is correct
corr_mat_melted = melt(corr_mat_lower)
corr_mat_melted$Var1 = factor(corr_mat_melted$Var1, levels=colnames(corr_mat))
corr_mat_melted$Var2 = factor(corr_mat_melted$Var2, levels=colnames(corr_mat))

# Plot
corr_plot = ggplot(corr_mat_melted, aes(x=Var1, y=Var2, fill=value)) +
    geom_tile(color='white') +
    scale_fill_distiller(limits=c(-1, 1), palette='RdBu', na.value='white',
                         name='Correlation') +
    ggtitle('Correlations') +
    coord_fixed(ratio=1) +
    theme_minimal() +
    scale_y_discrete(position="right") +
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          panel.grid.major=element_blank(),
          legend.position=c(0.1,0.9),
          legend.justification=c(0,1))
corr_plot
```






