---
title: "Untitled"
author: "Jesus"
date: "February 23, 2017"
output: html_document
runtime: shiny
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(class)
data(iris)
```


Boundary plot function from lab 6
```{r}
df = iris[which(iris$Species %in% c("versicolor", "virginica")),3:5]
resolution = 1000
  v1 = seq(min(df[,1]), max(df[,1]), length=resolution)
  v2 = seq(min(df[,2]), max(df[,2]), length=resolution)
  Grid = expand.grid(Petal.Length = v1, Petal.Width = v2)
```

```{r eruptions, echo=FALSE}
inputPanel(
  
  sliderInput("n_neigh", label = "Number of neighbors:", ticks = F,
              min = 1, max = 15, value = 1, step = 1)
)

renderPlot({
  irisknn <- knn(train = iris[,3:4], cl = iris$Species, test = Grid, k = as.numeric(input$n_neigh))
  Grid$Species = irisknn
  ggplot(data=df, aes(x=Petal.Length, y=Petal.Width, color = Species)) +
    geom_contour(data=Grid, aes(z=as.numeric(Species)),
                 color="black",size=0.5)+
    geom_point(size=2,aes(color=Species, shape=Species))
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



